The Multi-Stencil Framework: A Component-Based Approach to Orchestrate Stencils

### 2017-03-06 ARTICLES/IJPP/articleR1/formalism.tex
corrections formalism.tex + msl.tex

\paragraph{\textbf{Example}}
\begin{figure}[ht]
\begin{center}
\subfloat[On the left a mesh is represented, on the right two examples of groups of mesh entities are represented: cells and vertical edges.\label{fig:meshbase}]{
\resizebox{8cm}{!}{\includegraphics{./images/mesh.pdf}}
}\\
\hspace{10pt}
\subfloat[A is computed with a 4-neighborhood stencil applied on B. A is computed onto a computation domain which does not include all entities of the group.\label{fig:ex1}]{
\resizebox{5cm}{!}{\includegraphics{./images/stencil1.pdf}}
}
\vspace{20pt}
\hspace{10 pt}
\subfloat[A is computed with a 2-neighborhood stencil applied on C. A is computed onto a computation domain includes all entities of the group.\label{fig:ex2}]{
\resizebox{5cm}{!}{\includegraphics{./images/stencil2.pdf}}
}
\end{center}
\caption{(a) a Cartesian mesh and two kind of groups of mesh entities, (b) an example of stencil kernel on cells, (c) an example of stencil kernel on two different entities of the mesh.}
\label{fig:gspmsp}
\end{figure}
For example, in Figure~\ref{fig:ex1}, assuming that the computation domain (full lines) is denoted $dc1$ and the stencil shape described by the neighborhood function is $n1$, the stencil kernel can be defined as:
\begin{equation*}
R: \{(B,n1)\}, \quad w: A, \quad D: dc1,
\end{equation*}
\begin{equation*}
comp: A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
\end{equation*}
On the other hand, in the example of Figure~\ref{fig:ex2}, assuming the computation domain is $dc2$ and the stencil shape is $n2$, the stencil kernel is defined as:
\begin{equation*}
R: \{(C,n2),(A,identity)\}, \quad w: A, \quad D: dc2,
\end{equation*}
\begin{equation*}
comp: A(x,y)=A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

>>>>

\paragraph{\textbf{Example}}
\begin{figure}[ht]
\begin{center}
\subfloat[On the left a mesh is represented, on the right two examples of groups of mesh entities are represented: cells and vertical edges.\label{fig:meshbase}]{
\resizebox{8cm}{!}{\includegraphics{./images/mesh.pdf}}
}\\
\hspace{10pt}
\subfloat[A is computed with a 4-neighborhood stencil applied on B. A is computed onto a computation domain which does not include all entities of the group.\label{fig:ex1}]{
\resizebox{5cm}{!}{\includegraphics{./images/stencil1.pdf}}
}
\vspace{20pt}
\hspace{10 pt}
\subfloat[A is computed with a 2-neighborhood stencil applied on C. A is computed onto a computation domain includes all entities of the group.\label{fig:ex2}]{
\resizebox{5cm}{!}{\includegraphics{./images/stencil2.pdf}}
}
\end{center}
\caption{(a) a Cartesian mesh and two kind of groups of mesh entities, (b) an example of stencil kernel on cells, (c) an example of stencil kernel on two different groups of mesh entities.}
\label{fig:gspmsp}
\end{figure}
For example, in Figure~\ref{fig:ex1}, assuming that the computation domain (full lines) is denoted $dc1$ and the stencil shape described by the neighborhood function is $n1$, the stencil kernel can be defined as:
\begin{equation*}
R: \{(B,n1)\}, \quad w: A, \quad D: dc1,
\end{equation*}
\begin{equation*}
comp: A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
\end{equation*}
On the other hand, in the example of Figure~\ref{fig:ex2}, assuming the computation domain is $dc2$ and the stencil shape is $n2$, the stencil kernel is defined as:
\begin{equation*}
R: \{(C,n2),(A,identity)\}, \quad w: A, \quad D: dc2,
\end{equation*}
\begin{equation*}
comp: A(x,y)=A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

### 2017-03-06 ARTICLES/IJPP/articleR1/msl.tex
corrections formalism.tex + msl.tex

%----------------------------------------
\subsection{Generic Assembly}
%----------------------------------------
From the meta-model represented by Equation~(\ref{eq:msp}) the Generic Assembly (GA) of any multi-stencil program can be built. As illustrated in Figure~\ref{fig:ga}, the GA has five components: {\bf \emph{Driver}, \emph{Time}, \emph{DDS}, \emph{Data}, } and {\bf \emph{Computations}}.

@@

  \path[-]
    (Dp) edge node {} (D)
    (D) edge node {} (Du1)
        edge node {} (Du2)
        edge node {} (Du3)
    (DDSp1) edge node {} (DDS)
    (Tp) edge node {} (T)
    (T)  edge node {} (Tu)
    (Cp) edge node {} (C)
    (C) edge node {} (Cu)
    (Datap2) edge node {} (Data)
    (Datap1) edge node {} (Data)
    (Data) edge node {} (Datau)
          edge node {} (mpidata)
    (DDSp2) edge node {} (DDS);
\end{tikzpicture}
\vspace*{.5em}
\caption{Generic Assembly according to the Multi-Stencil program formalism.
Components circled by a double line identify those that will be instantiated multiple times at execution.
Component colors represent the actor of Figure~\ref{fig:msf} responsible for the component implementation: green for those automatically generated by the compiler, red for those implemented by HPC specialists and blue for those implemented by the developer.}
\label{fig:ga}
\end{center}
\end{figure}

\paragraph{\textbf{Driver}}
This component can be compared to the main function of a usual program. It is responsible for both the initialization and the execution of other components (like variable initializations and function calls).

\paragraph{\textbf{Time}}
This component is responsible for the time $T$ defined in Equation~(\ref{eq:msp}). It is composed of a time loop and potentially of a convergence reduction.

\paragraph{\textbf{DDS}}
This component is responsible for the mesh and its entities $\mathcal{M}$ and $\Phi$, the computation domains $\mathcal{D}$, and the set of neighborhood functions $\mathcal{N}$.
This component is generic; its type is a parameter of the generic assembly.
When the generic assembly is instantiated and specialized by MSC, a component handling a specific type of mesh is used in place of the DDS placeholder.
\CP{pas clair}
The interfaces exposed by this component are however well defined and any component providing these interfaces can be indifferently used.
Third party developers can therefore propose new DDS components.
In this paper, both data and task parallelism are used. In the case of data parallelism this component handles mesh partitioning and provides a synchronization interface as detailed in Section~\ref{sect:parallelism}.
The implementation of this component is the responsibility of HPC specialists.

\paragraph{\textbf{Data}}
This component is responsible for the set of mesh variables $\Delta$.
Each instance of the component uses the DDS component to handle one single mesh variable.
It is closely related to $DDS$ and its implementation is typically provided by the same HPC specialist as $DDS$.

\paragraph{\textbf{Computations}}
This component is responsible for $\Gamma$, \ie the computations of the simulation. 
It is automatically replaced by a sub-assembly of components produced by MSC for which the parallel part is automatically generated and the part responsible for the numerical code is filled by the developer.
The sub-assembly generation is described in Section~\ref{sect:parallelism}.

@@

\item The \texttt{mesh} keyword (Fig.~\ref{fig:grammar},~l.1) introduces an identifier for $\mathcal{M}$, the single mesh of the simulation. For example \texttt{cart} in Fig.~\ref{fig:exmsl},~l.1. The language, based on the meta-model is independent of the mesh topology, thus this identifier is actually not used by the compiler but is syntactic sugar for the user.
\JB{J'aime pas trop l'utilisation de ``syntactic sugar'' ici... Syntactic sugar est generalement utilise pour un racourci de syntaxe (i.e. int x = myInteger au lieu de int x = myInteger.getIntVal()), ce n'est pas le cas ici}

\item The \texttt{mesh entities} keyword (Fig.~\ref{fig:grammar},~l.2) introduces identifiers for the groups of entities $\mathcal{G}\in\Phi$. For example \texttt{cell} or \texttt{edgex} in Fig.~\ref{fig:exmsl},~l.2. Again, this is syntactic sugar only as the compiler does not rely on the mesh topology.

@@

\item The \texttt{stencil shapes} keyword (Fig.~\ref{fig:grammar},~l.5) introduces identifiers for $n\in\mathcal{N}$, the stencil shapes. For each shape, the source and destination group of entities (Fig.~\ref{fig:grammar},~l.16) are specified. For example \texttt{nec} in Fig.~\ref{fig:exmsl},~l.11 is a neighborhood from \texttt{edgex} to \texttt{cell}.

\item The \texttt{mesh quantities} keyword (Fig.~\ref{fig:grammar},~l.6) introduces identifiers for $\delta\in\Delta$, the mesh variables with the group of entities they are mapped on (Fig.~\ref{fig:grammar},~l.16). For example the quantities \texttt{C} and \texttt{H} are mapped onto \texttt{edgex} entities.

@@

\item Finally, the last part (Fig.~\ref{fig:grammar},~l.8) introduces the different computation loops of the simulation. Each loop is made of two parts:
\begin{itemize}
\item the \texttt{time} keyword (Fig.~\ref{fig:grammar},~l.22) introduces $conv$, the convergence criteria either as a constant number of iterations or as a scalar variable (Fig.~\ref{fig:grammar},~l.24). For example, 500 iterations are specified in Fig.~\ref{fig:exmsl},~l.16,
\item the \texttt{computations} keyword (Fig.~\ref{fig:grammar},~l.23) introduces identifiers for each computation $k\in\Gamma=(S,R,(w,D),comp)$. Each computation (Fig.~\ref{fig:grammar},~l.26) specifies:
\begin{itemize}
 \item the quantity $w$ written and its domain $D$, for example in Fig.~\ref{fig:exmsl},~l.22, kernel \texttt{k4} computes the quantity \texttt{F} on domain \texttt{d1},
 \item the read scalars $S$ and mesh quantities with their associated stencil shape $R$, for example in Fig.~\ref{fig:exmsl},~l.16, \texttt{k4} reads \texttt{C} with the shape \texttt{nce} and \texttt{D} with the default \textit{identity} shape; it does not read scalars.
 \end{itemize}
\end{itemize}
\end{enumerate}

One can notice that in the example of Figure~\ref{fig:exmsl}, there are no (reduction) kernel associated to the scalars \texttt{mu} and \texttt{tau}.
In this case, those scalars are in fact constants.
One can also notice that the computation to execute for each kernel is not specified. Only an identifier is given to each kernel, for example $k_4$. The numerical code is indeed not handled by MSL, which is mesh-agnostic, and which is used to generate a parallel orchestration of computations only. The numerical computation is specified after MSC compilation by the developer (Fig.~\ref{fig:msf}).

>>>>

From the octuplet of Equation~(\ref{eq:msp}) both the Generic Assembly (GA) of a multi-stencil program and the Multi-Stencil Language (MSL) can be built. GA and MSL are both are described in this section.

%----------------------------------------
\subsection{Generic Assembly}
%----------------------------------------
As illustrated in Figure~\ref{fig:ga}, the GA has five components: {\bf \emph{Driver}, \emph{Time}, \emph{DDS}, \emph{Data}, } and {\bf \emph{Computations}}. These components are generic components or abstract components. It means that interfaces of these components are well defined but that they are not implemented yet in GA. They can be compared to abstract classes and templates in \texttt{C++} for which an implementation must be given as well as specific parameters.

@@

  \path[-]
    (Dp) edge node {} (D)
    (D) edge node {} (Du1)
        edge node {} (Du2)
        edge node {} (Du3)
    (DDSp1) edge node {} (DDS)
    (Tp) edge node {} (T)
    (T)  edge node {} (Tu)
    (Cp) edge node {} (C)
    (C) edge node {} (Cu)
    (Datap2) edge node {} (Data)
    (Datap1) edge node {} (Data)
    (Data) edge node {} (Datau)
          edge node {} (mpidata)
    (DDSp2) edge node {} (DDS);
\end{tikzpicture}
\vspace*{.5em}
\caption{Generic Assembly according to the Multi-Stencil program formalism.
Components circled by a double line identify those that will be instantiated multiple times by MSC. Component colors represent actors of Figure~\ref{fig:msf} responsible for the component implementation: green for those automatically generated by the compiler, red for those implemented by HPC specialists and blue for those implemented by the developer.}
\label{fig:ga}
\end{center}
\end{figure}

\paragraph{\textbf{Driver}}
This component can be compared to the main function of a usual program. It is responsible for both the initialization and the execution of other components (like variable initialization and function calls). This component is generated by MSC (represented in green).

\paragraph{\textbf{Time}}
This component is responsible for the time $T$ defined in Equation~(\ref{eq:msp}). It is composed of a time loop and potentially of a convergence reduction. This component is generated by MSC (represented in green).

\paragraph{\textbf{DDS}}
This component is responsible for the mesh and its entities $\mathcal{M}$ and $\Phi$, the set of computation domains $\mathcal{D}$, and the set of neighborhood functions $\mathcal{N}$.
When the generic assembly is instantiated and specialized by MSC, an implementation of DDS is selected to handle a specific type of mesh.
%\CP{pas clair}
The interfaces exposed by this component are well defined and any component providing these interfaces can be indifferently used.
A third party specialist can therefore propose new implementation of DDS.
In this paper, both data and task parallelism are used. In the case of data parallelism DDS handles mesh partitioning and provides a synchronization interface as detailed in Section~\ref{sect:parallelism}. The implementation of this component is the responsibility of HPC specialists (represented in red).

\paragraph{\textbf{Data}}
This component is responsible for the set of mesh variables $\Delta$.
Each instance of the component uses the DDS component to handle one single mesh variable.
It is closely related to $DDS$ and its implementation is typically provided by the same HPC specialist as $DDS$ (represented in red).

\paragraph{\textbf{Computations}}
This component is responsible for $\Gamma$, \ie the computations of the simulation. 
It is automatically replaced by a sub-assembly of components produced by MSC for which the parallel part is automatically generated. On the other hand, components responsible for the numerical kernels are filled by the developer. This is why this component is represented in green and blue in Fig.~\ref{fig:ga}. The sub-assembly generation is described in Section~\ref{sect:parallelism}.

@@

\item The \texttt{mesh} keyword (Fig.~\ref{fig:grammar},~l.1) introduces an identifier for $\mathcal{M}$, the single mesh of the simulation. For example \texttt{cart} in Fig.~\ref{fig:exmsl},~l.1. The language, based on the meta-model is independent of the mesh topology, thus this identifier is actually not used by the compiler.% but is syntactic sugar for the user.
%\JB{J'aime pas trop l'utilisation de ``syntactic sugar'' ici... Syntactic sugar est generalement utilise pour un racourci de syntaxe (i.e. int x = myInteger au lieu de int x = myInteger.getIntVal()), ce n'est pas le cas ici}

\item The \texttt{mesh entities} keyword (Fig.~\ref{fig:grammar},~l.2) introduces identifiers for the groups of mesh entities $\mathcal{G}\in\Phi$. For example \texttt{cell} or \texttt{edgex} in Fig.~\ref{fig:exmsl},~l.2. %Again, this is syntactic sugar only as the compiler does not rely on the mesh topology.

@@

\item The \texttt{stencil shapes} keyword (Fig.~\ref{fig:grammar},~l.5) introduces identifiers for each stencil shape $n\in\mathcal{N}$. For each $n$, the source and destination group of entities (Fig.~\ref{fig:grammar},~l.16) are specified. For example \texttt{nec} in Fig.~\ref{fig:exmsl},~l.11 is a neighborhood from \texttt{edgex} to \texttt{cell}.

\item The \texttt{mesh quantities} keyword (Fig.~\ref{fig:grammar},~l.6) introduces identifiers for $\delta\in\Delta$, the mesh variables with the group of entities they are mapped on (Fig.~\ref{fig:grammar},~l.16). For example the quantities \texttt{C} and \texttt{H} are mapped onto the groups of mesh entities \texttt{edgex}.

@@

\item Finally, the last part (Fig.~\ref{fig:grammar},~l.8) introduces the different computation loops of the simulation. Each loop is made of two parts:
\begin{itemize}
\item the \texttt{time} keyword (Fig.~\ref{fig:grammar},~l.22) introduces either a constant number of iterations or $conv$, the convergence criteria that is a scalar (Fig.~\ref{fig:grammar},~l.24). For example, 500 iterations are specified in Fig.~\ref{fig:exmsl},~l.16,
\item the \texttt{computations} keyword (Fig.~\ref{fig:grammar},~l.23) introduces identifiers for each computation $k=(S,R,(w,D),comp) \in \Gamma$. Each computation (Fig.~\ref{fig:grammar},~l.26) specifies:
\begin{itemize}
 \item the quantity $w$ written and its domain $D$, for example in Fig.~\ref{fig:exmsl},~l.22, kernel \texttt{k4} computes the variable \texttt{F} on domain \texttt{d1},
 \item the read scalars $S$ and mesh variables with their associated stencil shape ($R$). For example in Fig.~\ref{fig:exmsl},~l.16, \texttt{k4} reads \texttt{C} with the shape \texttt{nce} and \texttt{D} with the default \textit{identity} shape; it does not read scalars.
 \end{itemize}
\end{itemize}
\end{enumerate}

One can notice that in the example of Figure~\ref{fig:exmsl}, there are no kernel associated to the scalars \texttt{mu} and \texttt{tau} (reduction).
In this case, those scalars are in fact constants.
One can also notice that the computation to execute for each kernel is not specified. Only an identifier is given to each kernel, for example $k_4$. The numerical code is indeed not handled by MSL that generates a parallel orchestration of computations only. The numerical computation is specified after MSC compilation by the developer (Fig.~\ref{fig:msf}).
